
---
title: "Quantium Virtual Internship - Retail Strategy and Analytics - Task 1"
mainfont: Roboto
monofont: Consolas
output:
 pdf_document: 
   df_print: default
   highlight: tango
   keep_tex: yes
   latex_engine: xelatex
header-includes:
  \usepackage{fvextra}
  \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
---

```{r setup, include=FALSE}
# set options for R markdown knitting
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(linewidth=80)
```
```{r knitr line wrap setup, include=FALSE}
# set up line wrapping in MD knit output
library(knitr)
hook_output = knit_hooks$get("output")
knit_hooks$set(output = function(x, options) 
{
 # this hook is used only when the linewidth option is not NULL
 if (!is.null(n <- options$linewidth)) 
 {
 x = knitr:::split_lines(x)
 # any lines wider than n should be wrapped
 if (any(nchar(x) > n)) 
 x = strwrap(x, width = n)
 x = paste(x, collapse = "\n")
 }
 hook_output(x, options)
})
```
# Solution template for Task 1
This file is a solution template for the Task 1 of the Quantium Virtual Internship.
It will walk you through the analysis, providing the scaffolding for your solution 
with gaps left for you to fill in yourself. 

Look for comments that say "over to you" for places where you need to add your own 
code! 
Often, there will be hints about what to do or what function to use in the text 
leading up to a code block - if you need a bit of extra help on how to use a 
function, the internet has many excellent resources on R coding, which you can find
using your favourite search engine. 
## Load required libraries and datasets
Note that you will need to install these libraries if you have never used these 
before. 
```{r 0 Load Libraries, results = 'hide'}
#### Example code to install packages
#install.packages("data.table")
#### Load required libraries
library("data.table")
library(ggplot2)
library(ggmosaic)
library(readr)
#### Point the filePath to where you have downloaded the datasets to and 
#### assign the data files to data.tables
# over to you! fill in the path to your working directory. If you are on a Windows machine, you will need to use forward slashes (/) instead of backshashes (\)
filePath <- "C:/Users/Luke Villanueva/source/repos/LukVill/Quantium Program/"
transactionData <- fread(paste0(filePath,"QVI_transaction_data.csv"))
customerData <- fread(paste0(filePath,"QVI_purchase_behaviour.csv"))
```
## Exploratory data analysis
The first step in any analysis is to first understand the data. Let's take a look 
at each of the datasets provided.
### Examining transaction data
We can use `str()` to look at the format of each column and see a sample of the 
data. As we have read in the dataset as a `data.table` object, we can also run 
`transactionData` in the console to see a sample of the data or use 
`head(transactionData)` to look at the first 10 rows.
Let's check if columns we would expect to be numeric are in numeric form and date 
columns are in date format.
```{r Examining transaction data}
#### Examine transaction data