
---
title: "Quantium Virtual Internship - Retail Strategy and Analytics - Task 1"
mainfont: Roboto
monofont: Consolas
output:
 pdf_document: 
   df_print: default
   highlight: tango
   keep_tex: yes
   latex_engine: xelatex
header-includes:
  \usepackage{fvextra}
  \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
---

```{r setup, include=FALSE}
# set options for R markdown knitting
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(linewidth=80)
```
```{r knitr line wrap setup, include=FALSE}
# set up line wrapping in MD knit output
library(knitr)
hook_output = knit_hooks$get("output")
knit_hooks$set(output = function(x, options) 
{
 # this hook is used only when the linewidth option is not NULL
 if (!is.null(n <- options$linewidth)) 
 {
 x = knitr:::split_lines(x)
 # any lines wider than n should be wrapped
 if (any(nchar(x) > n)) 
 x = strwrap(x, width = n)
 x = paste(x, collapse = "\n")
 }
 hook_output(x, options)
})
```
# Solution template for Task 1
This file is a solution template for the Task 1 of the Quantium Virtual Internship.
It will walk you through the analysis, providing the scaffolding for your solution 
with gaps left for you to fill in yourself. 

Look for comments that say "over to you" for places where you need to add your own 
code! 
Often, there will be hints about what to do or what function to use in the text 
leading up to a code block - if you need a bit of extra help on how to use a 
function, the internet has many excellent resources on R coding, which you can find
using your favourite search engine. 

## Load required libraries and datasets
Note that you will need to install these libraries if you have never used these 
before. 
```{r 0 Load Libraries, results = 'hide'}
#### Example code to install packages
#install.packages("data.table")
#### Load required libraries
library("data.table")
library(ggplot2)
library(ggmosaic)
library(readr)
library(plyr)

#### Point the filePath to where you have downloaded the datasets to and 
#### assign the data files to data.tables
# over to you! fill in the path to your working directory. If you are on a Windows machine, you will need to use forward slashes (/) instead of backshashes (\)
filePath <- paste0(getwd(),"/")
transactionData <- fread(file = paste0(filePath,"QVI_transaction_data.xlsx - in.csv"), header = TRUE)
customerData <- fread(file = paste0(filePath,"QVI_purchase_behaviour.csv"), header = TRUE)

```

## Exploratory data analysis
The first step in any analysis is to first understand the data. Let's take a look
at each of the datasets provided.
### Examining transaction data
We can use `str()` to look at the format of each column and see a sample of the
data. As we have read in the dataset as a `data.table` object, we can also run
`transactionData` in the console to see a sample of the data or use
`head(transactionData)` to look at the first 10 rows.
Let's check if columns we would expect to be numeric are in numeric form and date
columns are in date format.

```{r Examining transaction data}
#### Examine transaction data
# Over to you! Examine the data using one or more of the methods described above.

str(transactionData)

transactionData

```

We can see that the date column is in an integer format. Let's change this to a
date format.

```{r Convert DATE to date format}
#### Convert DATE column to a date format
#### A quick search online tells us that CSV and Excel integer dates begin on 30 Dec 1899

transactionData$DATE <- as.Date(transactionData$DATE, origin = "1899-12-30")

```

We should check that we are looking at the right products by examining PROD_NAME.

```{r Summary of PROD_NAME}
#### Examine PROD_NAME
# Over to you! Generate a summary of the PROD_NAME column.

transactionData$PROD_NAME[1:10]

summary(transactionData$PROD_NAME)

```

Looks like we are definitely looking at potato chips but how can we check that
these are all chips? We can do some basic text analysis by summarizing the
individual words in the product name.

```{r Further examine PROD_NAME}
#### Examine the words in PROD_NAME to see if there are any incorrect entries
#### such as products that are not chips

# make dataframe of all chars in all data names
productWords <- data.table(unlist(strsplit(unique(transactionData[, PROD_NAME]), "")))

# set column name as words
setnames(productWords, 'words')
```

As we are only interested in words that will tell us if the product is chips or
not, let's remove all words with digits and special characters such as '&' from our
set of product words. We can do this using `grepl()`.

```{r}
# Over to you! Remove digits, and special characters, and then sort the distinct words by frequency of occurrence.

#### Removing digits
productWords <- productWords[!grepl("\\d",productWords$words)]


#### Removing special characters
# remove punct
productWords <- productWords[!grepl("[[:punct:]]",productWords$words)]

#### Let's look at the most common words by counting the number of times a word appears and
#### sorting them by this frequency in order of highest to lowest frequency

# make vector of each word by extracting each sequence of letters surrounded by spaces
# for each entry in vector char, if entry is space, then everything before it is a word
word_vec <- character(0)
beg_indx <- 1
for(i in seq_along(productWords$words))
{
  # if space
  if (productWords$words[i] == " ")
  {
    # make all characters appended to word
    str_word <- ""
    for(j in (seq_len(i - beg_indx) + beg_indx - 1))
    {
      str_word <- paste0(str_word,productWords$words[j])
    }
    
    # add word to word_vec
    word_vec <- c(word_vec,str_word)
    
    # afterward, change beginning index to index after i
    beg_indx <- i + 1
  }
  
  # else, increment counter
}

# with word_vec, count most common words, exclude ""
summary_word_vec <- sort(table(word_vec))
summary_word_vec <- rev(summary_word_vec[-length(summary_word_vec )])

# print out sorted vector
summary_word_vec

```

There are salsa products in the dataset but we are only interested in the chips
category, so let's remove these.

```{r}
#### Remove salsa products
transactionData[, SALSA := grepl("salsa", tolower(PROD_NAME))]
transactionData <- transactionData[SALSA == FALSE, ][, SALSA := NULL]
```

Next, we can use `summary()` to check summary statistics such as mean, min and max
values for each feature to see if there are any obvious outliers in the data and if
there are any nulls in any of the columns (`NA's : number of nulls`) will appear in
the output if there are any nulls).

```{r initial summary}
#### Summarise the data to check for nulls and possible outliers
# Over to you!

summary(transactionData)

```

There are no nulls in the columns but product quantity appears to have an outlier
which we should investigate further. Let's investigate further the case where 200
packets of chips are bought in one transaction.

```{r }
#### Filter the dataset to find the outlier
# Over to you! Use a filter to examine the transactions in question.

transactionData[transactionData$PROD_QTY == 200.000]

```

There are two transactions where 200 packets of chips are bought in one transaction
and both of these transactions were by the same customer.

```{r}
#### Let's see if the customer has had other transactions
# Over to you! Use a filter to see what other transactions that customer made.

transactionData[transactionData$LYLTY_CARD_NBR == 226000]

```

It looks like this customer has only had the two transactions over the year and is
not an ordinary retail customer. The customer might be buying chips for commercial
purposes instead. We'll remove this loyalty card number from further analysis.

```{r}
#### Filter out the customer based on the loyalty card number
# Over to you!

transactionData <- transactionData[!transactionData$LYLTY_CARD_NBR == 226000]

#### Re-examine transaction data
# Over to you!

summary(transactionData)

```

That's better. Now, let's look at the number of transaction lines over time to see
if there are any obvious data issues such as missing data.

```{r}
#### Count the number of transactions by date
# Over to you! Create a summary of transaction count by date.

# get vec of dates in data, for each entry in data
count_trans <- numeric(0)
for(i in seq_along(unique(transactionData$DATE)))
{
  # filter by each date, get row count, then place count in result vec
  count_trans[i] <- nrow(transactionData[transactionData$DATE == unique(transactionData$DATE)[i]])
}

# rename names of count_trans
names(count_trans) <- unique(transactionData$DATE)

# sort transaction dates
names(count_trans) <- sort(names(count_trans))

# get number of dates
length(count_trans)

```

There's only 364 rows, meaning only 364 dates which indicates a missing date. Let's
create a sequence of dates from 1 Jul 2018 to 30 Jun 2019 and use this to create a
chart of number of transactions over time to find the missing date.

```{r fig.align = "center"}
#### Create a sequence of dates and join this the count of transactions by date
# Over to you - create a column of dates that includes every day from 1 Jul 2018 to 30 Jun 2019, and join it onto the data to fill in the missing day.

dates_seq <- character(0)

for(i in seq_len(365))
{
  dates_seq <- c(dates_seq, format(as.Date(i, origin = "2018-06-30"), "%Y-%m-%d"))
}



#### Setting plot themes to format graphs
theme_set(theme_bw())
theme_update(plot.title = element_text(hjust = 0.5))

#### Plot transactions over time
ggplot(transactions_by_day, aes(x = DATE, y = N)) +
 geom_line() +
 labs(x = "Day", y = "Number of transactions", title = "Transactions over time") +
 scale_x_date(breaks = "1 month") +
 theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

```