
---
title: "Quantium Virtual Internship - Retail Strategy and Analytics - Task 1"
mainfont: Roboto
monofont: Consolas
output:
 pdf_document: 
   df_print: default
   highlight: tango
   keep_tex: yes
   latex_engine: xelatex
header-includes:
  \usepackage{fvextra}
  \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
---

```{r setup, include=FALSE}
# set options for R markdown knitting
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(linewidth=80)
```
```{r knitr line wrap setup, include=FALSE}
# set up line wrapping in MD knit output
library(knitr)
hook_output = knit_hooks$get("output")
knit_hooks$set(output = function(x, options) 
{
 # this hook is used only when the linewidth option is not NULL
 if (!is.null(n <- options$linewidth)) 
 {
 x = knitr:::split_lines(x)
 # any lines wider than n should be wrapped
 if (any(nchar(x) > n)) 
 x = strwrap(x, width = n)
 x = paste(x, collapse = "\n")
 }
 hook_output(x, options)
})
```
# Solution template for Task 1
This file is a solution template for the Task 1 of the Quantium Virtual Internship.
It will walk you through the analysis, providing the scaffolding for your solution 
with gaps left for you to fill in yourself. 

Look for comments that say "over to you" for places where you need to add your own 
code! 
Often, there will be hints about what to do or what function to use in the text 
leading up to a code block - if you need a bit of extra help on how to use a 
function, the internet has many excellent resources on R coding, which you can find
using your favourite search engine. 

## Load required libraries and datasets
Note that you will need to install these libraries if you have never used these 
before. 
```{r 0 Load Libraries, results = 'hide'}
#### Example code to install packages
#install.packages("data.table")
#### Load required libraries
library("data.table")
library(ggplot2)
library(ggmosaic)
library(readr)
#### Point the filePath to where you have downloaded the datasets to and 
#### assign the data files to data.tables
# over to you! fill in the path to your working directory. If you are on a Windows machine, you will need to use forward slashes (/) instead of backshashes (\)
filePath <- paste0(getwd(),"/")
transactionData <- fread(file = paste0(filePath,"QVI_transaction_data.xlsx - in.csv"), header = TRUE)
customerData <- fread(file = paste0(filePath,"QVI_purchase_behaviour.csv"), header = TRUE)

```

## Exploratory data analysis
The first step in any analysis is to first understand the data. Let's take a look
at each of the datasets provided.
### Examining transaction data
We can use `str()` to look at the format of each column and see a sample of the
data. As we have read in the dataset as a `data.table` object, we can also run
`transactionData` in the console to see a sample of the data or use
`head(transactionData)` to look at the first 10 rows.
Let's check if columns we would expect to be numeric are in numeric form and date
columns are in date format.

```{r Examining transaction data}
#### Examine transaction data
# Over to you! Examine the data using one or more of the methods described above.

str(transactionData)

transactionData

```

We can see that the date column is in an integer format. Let's change this to a
date format.

```{r Convert DATE to date format}
#### Convert DATE column to a date format
#### A quick search online tells us that CSV and Excel integer dates begin on 30 Dec 1899

transactionData$DATE <- as.Date(transactionData$DATE, origin = "1899-12-30")
```

We should check that we are looking at the right products by examining PROD_NAME.

```{r Summary of PROD_NAME}
#### Examine PROD_NAME
# Over to you! Generate a summary of the PROD_NAME column.

transactionData$PROD_NAME[1:10]

summary(transactionData$PROD_NAME)

```

Looks like we are definitely looking at potato chips but how can we check that
these are all chips? We can do some basic text analysis by summarizing the
individual words in the product name.

```{r Further examine PROD_NAME}
#### Examine the words in PROD_NAME to see if there are any incorrect entries
#### such as products that are not chips

# make dataframe of all chars in all data names
productWords <- data.table(unlist(strsplit(unique(transactionData[, PROD_NAME]), "")))

# set column name as words
setnames(productWords, 'words')
```

As we are only interested in words that will tell us if the product is chips or
not, let's remove all words with digits and special characters such as '&' from our
set of product words. We can do this using `grepl()`.

```{r}
# Over to you! Remove digits, and special characters, and then sort the distinct words by frequency of occurrence.

#### Removing digits
productWords <- productWords[!grepl("\\d",productWords$words)]


#### Removing special characters
productWords <- productWords[!grepl("[$&+,:;=?@#|'<>.^*()%!]-/",productWords$words)]

#### Let's look at the most common words by counting the number of times a word appears and
#### sorting them by this frequency in order of highest to lowest frequency

# make vector of each word by extracting each sequence of letters surrounded by spaces
# for each entry in vector char, if entry is space, then everything before it is a word
word_vec <- character(0)
beg_indx <- 1
for(i in seq_along(productWords$words))
{
  # if space
  if (productWords$words[i] == " ")
  {
    # make all characters appended to word
    str_word <- ""
    for(j in (seq_len(i - beg_indx) + beg_indx - 1))
    {
      str_word <- paste0(str_word,productWords$words[j])
    }
    
    # add word to word_vec
    word_vec <- c(word_vec,str_word)
    
    # afterward, change beginning index to index after i
    beg_indx <- i + 1
  }
  
  # else, increment counter
}

# with word_vec, count most common words, exclude ""
summary_word_vec <- sort(table(word_vec))
summary_word_vec <- summary_word_vec[-length(summary_word_vec )]

```
